### 2024-11-07 MsaPythonDataAnalytics.qmd
update values

---
title: "Data Analytics Introduction Using Python Training"
jupyter: python3
format: 
  revealjs:
    theme: serif
    smaller: false
    scrollable: true
---

# Introduction

## Welcome

We are embarking on a journey, with multiple stops but the destination will be far out. The stops will be areas of expertise and learning that we will have done and practiced but the destination is the progressive accumulation of the learning.

With the ever evolving changing of technology the destination will always be changing.

------------------------------------------------------------------------

::: {.columns}

::: {.column width="60%"}
### Assumptions {.smaller}

Have some limited or little knowledge of Microsoft excel functions e.g `=sum(C1,C2)`, though not mandatory it makes it easier to grasp concepts slightly faster. Though if you remember the simple mathematics we do everyday of summing and adding money then you are good :).
:::

::: {.column width="40%"}
### Requirements {.smaller}

-   *Interest* and *Desire* to learn
-   Working Computer (Laptop/Desktop)
-   Access to internet once in a while to download a few resource materials
:::
:::

------------------------------------------------------------------------

::: {.columns}

::: {.column width="40%"}
### Teaching guide

-   Mixed instructional guide, with the presentation but focused on getting more hard skills

-   Engagements and practice on the fly
:::

::: {.column width="60%"}
### About

Just me but you can see you can get the details from [qprop](https://www.qprop.me/about/)

Will be getting assisted by Alex.
:::
:::

## Data Analytics {.smaller}

Colloquial term ***data analytics*** can be coined as both a science and an art. With the ***science*** part majorly following statistical/mathematical procedures used, *art* comes from the different ways and methods someone can use to present and execute the since part of it.

### Tools for Data Analytics

There are various tools/Software/applications can be used for data analysis

::: {.columns}

::: {.column width="50%"}
- Paid for: 
  - Ms Excel
  - Ms Power BI
  - SPSS
  - STATA
  - SAS
  - MS SQL

:::
::: {.column width="50%"}

- Open Source: **Free**
  - ***Python***
  - R
  - Postgres
  - Julia
  - MangoDB
  - CouchDB

:::
:::

# Python

## Python Introduction  {.smaller}

Open Source Programming language.First come to use in the early 1990's and developed by **Guido Van Rossum** more information [here](https://en.wikipedia.org/wiki/Python_(programming_language)).

::: {.columns}

::: {.column width="50%"}

### Advantages

- Easy to learn
- Almost language like syntax
- Fast execution
- All Purpose programming language:
  - used for software development
  - used for data analysis
  - used for machine learning
  - used for web development

:::

::: {.column width="50%"}

### Disadvantages

- Some convections are different from other programming language
- High level interpreted language

:::
:::

## Python Interface

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

```{python}
#| echo: true

1 + 1
```

## Python modules

## Python numpy | Introduction

## Python Pandas | Introduction

## Python Visualization | Matplotlip & Seaborn

## Folder Structure & File Management

# SQL

## SQL Introduction  {.smaller}

### 2024-11-07 MsaPythonDataAnalytics.qmd
update values 1

@@
     scrollable: true
+code-fold: true
 ---
@@
 
-# Python
+# Python {.smaller}
 
-## Python Introduction  {.smaller}
+## Python Introduction
 
@@
 
+Python programming language uses it's in build command prompt frequently called *CLI* standing for *Command Line Interface*, search through windows/MAC program files and you should see *Python 3.0.0)* there are various versions of python and depending on which one you have installed this will determine the python CLI.
+
+Though you can be able to do everything using this, it doesn't give an intuitive user interface hence the reason for development of *IDE* Integrated Development Environment.
+
+*IDE* is the dashboard similar to car dashboard but the actual software the engine is now *python* for this case.
+
 Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
@@
 
-## Python numpy | Introduction
+## Python numpy | Introduction {.smaller}
+
+***numpy***: python module/library specialized in *Arrays and Vectorized Computation*.
+
+NumPy, short for Numerical Python, is one of the most important foundational packages for numerical computing in Python. Numpy works with array from 1-n dimenisional.
+
+```{python}
+
+#| echo: true
+
+import numpy as np
+
+my_arr = np.arange(10)
+
+#my_list = list(range(10)) #inbuilt python
+
+my_arr
+```
+
+Multidimensional Array
+
+```{python}
+#| echo: true
+
+data = np.array([[1.5, -0.1, 3], [0, -3, 6.5]])
+
+data
+
+```
+
+Array descriptors
+
+```{python}
+#| echo: true
+
+data.shape
+data.dtype
+data.ndim
+
+```
 
 ## Python Pandas | Introduction
+
+***Pandas*** : Python module/library enhancing data manipulation tools designed to make data cleaning and analysis fast and convenient in Python.
+Works intandem with *numpy* and is the core working under the hood.
+
+```{python}
+#| echo: true
+
+import pandas as pd
+
+obj_series = pd.Series([4, 7, -5, 3])
+
+obj_series
+
+```
+
+Main difference of pandas with numpy is that it has indexed values and designed for working with tabular or heterogeneous data.
 

### 2024-11-08 MsaPythonDataAnalytics.qmd
content day 1

@@
     theme: serif
-    smaller: false
+    smaller: true
     scrollable: true
@@
 
-## Python Interface
+## Python Interface {.smaller}
 
@@
 
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
+::: {.panel-tabset}
+
+### IDE's Common to Python
+
+- Pycharm
+- Spyder
+- vsCode
+- Rstudio
+- Positron (*New in Beta*)
+- Text editors
+  - notepad ++
+  - vim
+  - sublime
+- e.t.c
+
+Though there are many and the tool of choice is open for use, for now we can focus on using vsCode as this is universal to also other programming languages but if you are interested in a polygot system you can test positron.
+
+### Installations
+
+***Let us check what installations you have***
+
+- Python Installation
+- IDE Installation
+- To install all this together we prefer [anaconda](www.anaconda.com)
+  - Anaconda | **full suit of packages and tools**
+  - miniconda | **minimal and necessary packages**
+
+:::
+
+## Python | Hello World
+
+Python being an interpreted high level programming language, making things easier for the programmer. Able to pick up the things very easily.
+
+Let us start with the first code.
+
+::: {.panel-tabset}
+
+### Hello World
+
+```{python}
+#| echo: True
+
+print("Hello WOrld")
+
+```
+
+Let us use it as a calculator.
+
+```{python code-line-numbers="4,6,8,10"}
+#| echo: true
+#| eval: false
+
+2 + 2  #Add
+#4
+3 - 1  #Subtract
+#2
+4 * 5  #Multiply
+#20
+20 / 5 #Divide
+#4 
+5 ** 2 #Exponent
+#25
+5 % 2  #?What is the result
+
+```
+
+Using it as an ***input***
+
+```{python}
+#| echo: true
+#| eval: false
+
+input("What's your name")
+
+```
+
+Using ***comments***
 
@@
 
-1 + 1
+#This is a comment
+
 ```
+
+### Data types
+
+**Strings**
+
+```{python}
+#| echo: true
+first_string = "My first String"
+first_string
+```
+
+***List***
+
+```{python}
+#| echo: true
+first_list = [1,2,3,4]
+first_list
+```
+
+***Tuples***
+```{python}
+#| echo: true
+first_tuple = (1,2,3)
+first_tuple
+```
+
+***Dictionary***
+```{python}
+#| echo: true
+first_dic = {'a' : [1,2,3], 'b' : [4,5,6], 'c' : [7,6,8]}
+first_dic
+```
+
+### Methods & Functions
+
+**Methods**
+
+Method: Attributes associated to different objects and data types. As well classes at a broader level
+
+```{python}
+#| echo: true
+first_string.upper()
+first_string.lower()
+first_string.swapcase()
+first_string.rsplit() #separating or delimiter is a space
+
+```
+
+***Functions***
+
+```{python}
+#| echo: true
+first_list = [1,2,3,4]
+first_list
+```
+
+### Summary Sheets
+
+|**Operator**|**Description**|**Associativity**|
+|----|-----|------|
+|`{key:expr,...}`   |Dictionary creation| NA|
+|`{ expr ,...}`     |Set creation       | NA|
+|`[ expr ,...]`     |List creation      | NA|
+|`( expr ,...)`     |Tuple creation or just parentheses      | NA|
+|`f ( expr ,...)`     |Function call      | L|
+|`x [ index : index ]`     |Slicing      | L|
+|`x [ index ]`     |Indexing      | L|
+|`x . attr`     |Attribute reference      | L|
+|`x ** y`     |Exponentiation (x to the yth power)      | R|
+|`~ x`     |Bitwise NOT      | NA|
+|`+x, -x`     |Unary plus and minus      | NA|
+|`x*y, x/y, x//y, x%y`     |Multiplication, division, truncating division,remainder      | L|
+|`x+y, x-y` | Addition, subtraction | L |
+|`x<<y, x>>y` | Left-shift, right-shift | L |
+|`x & y` | Bitwise AND | L |
+|`x ^ y` | Bitwise XOR | L |
+|`x | y` | Bitwise OR | L |
+|`x<y, x<=y, x>y, x>=y, x<>y (v2 only),x!=y, x==y` | Comparisons (less than, less than or equal, greater than, greater than or equal, inequality, equality)a | NA |
+|`x is y, x is not y` | Identity tests | NA |
+|`x in y, x not in y` | Membership tests | NA |
+|`not x` | Boolean NOT | NA | 
+|`x and y` | Boolean AND | L | 
+|`x or y` | Boolean OR | L | 
+|`x if expr else y` | Ternary operator | NA |
+|`lambda arg,...: expr` | Anonymous simple function | NA |
+
+:::
 
@@
 
-NumPy, short for Numerical Python, is one of the most important foundational packages for numerical computing in Python. Numpy works with array from 1-n dimenisional.
+NumPy, short for Numerical Python, is one of the most important foundational packages for numerical computing in Python. Numpy works with array from 1-n dimensional.
 

### 2024-11-08 MsaPythonDataAnalytics.qmd
'updates'

@@
     scrollable: true
+    incremental: true
+    transition: zoom
+    background-transition: fade
 code-fold: true
@@
 
-```{python code-line-numbers="4,6,8,10"}
+```{python}
 #| echo: true
 #| eval: false
+#| code-line-numbers: "|1|3|5|7"
 
@@
 
+### Keywords
+
+**python** has *keywords* this are words that have syntactical use in the program below list even though not fully conclusive.
+
+`and`   `continue`    `except`    `global`    `lambda`    `raise`   `yield`
+
+`as`    `def`   `exec`    `if`    `not`   `return`
+
+`assert`    `del`   `finally`   `import`    `or`    `try`
+
+`break`   `elif`    `for`   `in`    `pass`    `while`
+
+`class`   `else`    `from`    `is`    `print`   `with`
+
 ### Data types
 
+**Numbers**
+**Integer Numbers**
+
 **Strings**
+
+sequence of characters used to store and represent text-based information
 
@@
 
+```{python}
+#| echo: true
+#| eval: false
+long_string = """Very long string
+spanning multiple lines
+that never seem to come to an end"""
+long_string
+```
+
 ***List***
+
+Mutable ordered sequence of items.
 
@@
 #| echo: true
-first_list = [1,2,3,4]
+first_list = [1,2,3,4,1,1,1,1]
 first_list
@@
 
+List objects provide several methods
+
+```{python}
+#| echo: true
+#| eval: false
+first_list.count(1)
+first_list.index(1)
+first_list.append(5)
+first_list.remove(5)
+first_list.pop(-1)
+first_list.reverse()
+first_list.sort()
+```
+
 ***Tuples***
+
+Immutable ordered sequence of items.
+
 ```{python}
@@
 
+***Sets***
+
+Ordered collections of unique items.
+
+```{python}
+#| echo: true
+#| #| eval: false
+{42, 3.14, 'hello'} # Literal for a set with three items
+{100} # Literal for a set with one item
+set() # Empty set (can't use {}—empty dict!)
+```
+
 ***Dictionary***
+
+Arbitrary collection of objects indexed by nearly arbitrary values called keys.
+
 ```{python}
@@
 first_dic
+```
+
+```{python}
+#| echo: true
+#| eval: false
+
+{'x':42, 'y':3.14, 'z':7} # Dictionary with three items, str keys
+{1:2, 3:4} # Dictionary with two items, int keys
+{1:'za', 'br':23} # Dictionary with mixed key types
+{} # Empty dictionary
+
+dict(x=42, y=3.14, z=7) # Dictionary with three items, str keys
+dict([(1, 2), (3, 4)]) # Dictionary with two items, int keys
+dict([(1,'za'), ('br',23)]) # Dictionary with mixed key types
+dict() # Empty dictionary
 ```
@@
 ### Summary Sheets
+
+**associativity** of the operator: L (left-to-right), R (right-to-left), or NA (nonassociative).
 

### 2024-11-08 MsaPythonDataAnalytics.qmd
update 2

@@
     scrollable: true
-    incremental: true
-    transition: zoom
+    incremental: false
+    transition: concave
     background-transition: fade
+    controls: true
     code-fold: true
+    code-tools: true
+
 ---
@@
 
+Assigning objects to names **variables**
+
+```{python}
+#| echo: true
+
+department = "DT"
+
+print(department)
+
+# you can change the variable on the fly
+
+department = 'Customs'
+
+print(department)
+```
+
 ### Keywords
@@
 **Numbers**
-**Integer Numbers**
+
+Integers, floating point numbers and complex numbers falls under Python numbers category.
+
+> We can use the `type()` function to know which class a variable or a value belongs to and the `isinstance()` function to check if an object belongs to a particular class.
+
+```{python}
+#| echo: true
+
+a = 5
+print(a, "is of type", type(a))
+
+a = 2.0
+print(a, "is of type", type(a))
+
+a = 1+2j
+print(a, "is of type", type(a))
+print(a, "is complex number?", isinstance(1+2j,complex))
+
+```
 
@@
 
-Immutable ordered sequence of items.
+Immutable ordered sequence of items.Tuples once created cannot be modified.
 
@@
 
+## Modules introduction
+
+When our program grows bigger, it is a good idea to break it into different modules.
+
+A module is a file containing Python definitions and statements. Python modules have a filename and end with the extension .py.
+
+Definitions inside a module can be imported to another module or the interactive interpreter in Python. We use the import keyword to do this.
+
+For example, we can import the math module by typing in import math.
+
+```{python}
+#| echo: true
+
+import math
+print(math.pi)
+
+```
+
+Checking paths using **sys** module
+
+```{python}
+#| echo: true
+
+import sys
+print(sys.path)
+```
+
 ## Python numpy | Introduction {.smaller}
@@
 
+::: {.panel-tabset}
+
+### Matplotlip
+
+Matplotlib is the "grandfather" library of data visualization with Python.
+It was created by John Hunter.
+He created it to try to replicate MatLab's (another programming language) plotting capabilities in Python.
+So if you happen to be familiar with matlab, matplotlib will feel natural to you.
+
+It is an excellent 2D and 3D graphics library for generating scientific figures.
+
+Some of the major Pros of Matplotlib are:
+
+* Generally easy to get started for simple plots
+* Support for custom labels and texts
+* Great control of every element in a figure
+* High-quality output in many formats
+* Very customizable in general
+
+Matplotlib allows you to create reproducible figures programmatically.
+Let's learn how to use it!
+Before continuing this lecture, I encourage you just to explore the official Matplotlib web page: [http://matplotlib.org/](http://matplotlib.org/)
+
+```{python}
+#| echo: true
+#| eval: true
+import matplotlib.pyplot as plt
+```
+
+```{python}
+#| echo: true
+import numpy as np
+x = np.linspace(0, 5, 11)
+y = x ** 2
+```
+
+Since we are using ipython notebook we need to run `%matplotlib inline` before running the plot
+
+```{python}
+#| echo: true
+plt.plot(x, y, 'r') # 'r' is the color red
+plt.xlabel('X Axis Title Here')
+plt.ylabel('Y Axis Title Here')
+plt.title('String Title Here')
+plt.show()
+```
+
+### Seaborn
+
+:::
+
 ## Folder Structure & File Management

### 2024-11-09 MsaPythonDataAnalytics.qmd
'updates_day3''

@@
 - Spyder
+- *jupyter*
 - vsCode
@@
 ```
+
+Rules for variable names
+
+* Can't start with numbers (1,2,3,4....)
+* Letter, numbers, underscores are allowed in the name but ',-, spaces are not allowed
 
@@
 
+## Python | Further into Hello World
+
+::: {.panel-tabset}
+
+### Iterations
+
+Iterations is repeating and the most common form of iteration is `for` 
+Code highligt for `for`
+
+```{python}
+#| echo: false
+#| eval: false
+
+for value in a_list:
+  #do something
+
+```
+
+:::
+
 ## Python modules
@@
 
+::: {.panel-tabset}
+
+### Intro
+
+Pandas relies on dataframes this is excel like data format with rows/records and columns/fields. Which mankes it easy to work with.
+
+Each row stands for an observation and columns here are variables.
+
+```{python}
+#| echo: true
+
+data = {
+  'county' : ['Nairobi','Kiambu','Kajiado','Machakos'],
+  'headquarters' : ['Nairobi','Kiambu','Kajiado','Machakos'],
+  'population' : [4397073,2417735,1117840,1421932]
+}
+
+df_data = pd.DataFrame(data,columns=['county','headquarters','population'])
+df_data
+```
+
+Checking and investigating the dataframe
+
+```{python}
+#| echo: true
+#| eval: false
+
+df_data.shape()
+df_data.index()
+df_data.columns()
+df_data.info()
+df_data.count()
+
+```
+
+### Pandas Summary
+
+```{python}
+#| echo: true
+#| eval: false
+
+df_data.sum()
+df_data.cumsum()
+df_data.min()
+df_data.max()
+df_data.idmax()
+df_data.idmin()
+df_data.describe()
+df_data.mean()
+df_data.media()
+```
+
+### Pandas Import/Export [Read/Write]
+
+**Read CSV**
+
+```{python}
+#| echo: false
+#| eval: false
+
+#import pandas as pd
+df = pd.read_csv('file.csv', header = None, nrows=5)
+df.to_csv("first_dataframe.csv")
+
+```
+
+**Read excel**
+
+```{python}
+#| echo: false
+#| eval: false
+
+df = pd.read_excel('excel_file.xlsx', sheet = 'Sheet1')
+
+df_mulitple_excel = pd.ExcelFile('excel_file.xlsx')
+df = pd.read_excel(df_mulitple_excel,  'Sheet1')
+
+df.to_excel('first_dataframe.xlsx', sheet_name = 'Sheet first')
+```
+
+|Function|	Description
+|-------|--------
+|read_csv |	Load delimited data from a file, URL, or file-like object; use comma as default delimiter
+|read_fwf |	Read data in fixed-width column format (i.e., no delimiters)
+|read_clipboard|	Variation of `read_csv` that reads data from the clipboard; useful for converting tables from web pages
+|read_excel|	Read tabular data from an Excel XLS or XLSX file
+|read_hdf|	Read HDF5 files written by pandas
+|read_html|	Read all tables found in the given HTML document
+|read_json|	Read data from a JSON (JavaScript Object Notation) string representation, file, URL, or file-like object
+|read_feather|	Read the Feather binary file format
+|read_orc|	Read the Apache ORC binary file format
+|read_parquet|	Read the Apache Parquet binary file format
+|read_pickle|	Read an object stored by pandas using the Python pickle format
+|read_sas|	Read a SAS dataset stored in one of the SAS system's custom storage formats
+|read_spss|	Read a data file created by SPSS
+|read_sql|	Read the results of a SQL query (using SQLAlchemy)
+|read_sql_table|	Read a whole SQL table (using SQLAlchemy); equivalent to using a query that selects everything in that table using `read_sql`
+|read_stata|	Read a dataset from Stata file format
+|read_xml|	Read a table of data from an XML file
+
+:::
+
 ## Python Visualization | Matplotlip & Seaborn

### 2024-11-10 MsaPythonDataAnalytics.qmd
update 4

@@
 
--   Mixed instructional guide, with the presentation but focused on getting more hard skills
+>   Mixed instructional guide, with the presentation but focused on getting more hard skills
 
--   Engagements and practice on the fly
+>   Engagements and practice on the fly
 :::
@@
 ```{python}
-#| echo: false
+#| echo: true
 #| eval: false
@@
 
+### Pandas Data Cleaning
+
+> 80% of the work done on data is cleaning
+
+#### Dealing with missing data
+
+|Method |	Description
+|------|---------
+|dropna |	Filter axis labels based on whether values for each label have missing data, with varying thresholds for how much missing data to tolerate.
+|fillna |	Fill in missing data with some value or using an interpolation method such as "ffill" or "bfill".
+|isna |	Return Boolean values indicating which values are missing/NA.
+|notna	| Negation of `isna`, returns True for non-NA values and False for NA values.
+
+#### Data Transformation
+
+***Removing Duplicates***
+
+```{python}
+#| echo: true
+#| eval: false
+
+data = pd.DataFrame({"k1": ["one", "two"] * 3 + ["two"],
+                     "k2": [1, 1, 2, 3, 3, 4, 4]})
+
+data
+
+data.duplicated()
+
+data.drop_duplicates()
+
+data["v1"] = range(7)
+
+data
+
+data.drop_duplicates(subset=["k1"])
+
+data.drop_duplicates(["k1", "k2"], keep="last")
+
+```
+
 :::
+
+## Visualization
+
+The point of visualization is being able to summary data to palatable chunks, easy to read and digest. Quick glances and be able to tell a story.
+
+Humans are visual, absorbing more through image than any other thing. There are very many examples of visuals that have impacted the world I will talk about 3.
+
+1. **The general** [minrad maps](https://bigthink.com/wp-content/uploads/2010/07/18578322.png?fit=1200,675)
+2. **The nurse** [crimea wars](https://upload.wikimedia.org/wikipedia/commons/1/17/Nightingale-mortality.jpg)
+3. **The Doctor** [cholera maps](https://www.esri.com/arcgis-blog/wp-content/uploads/2023/12/JohnSnowCholeraMap_Detail.png)
+
+Any visual should be able to tell a story either for record purposes and future solutions, predictions.
+
+## {fullscreen=true}
+
+<iframe class="stretch" data-src="https://www.gapminder.org/tools/#$chart-type=bubbles&url=v1"></iframe>
 
@@
 
+**Bar charts**
+
+```{python}
+#| echo: true
+x = [10, 20, 30, 40, 50, 60]
+y = [13, 45, 23, 34, 96, 76]
+plt.title('Bar Graph')
+plt.bar(x, y, color='dodgerblue', width=5)
+plt.show()
+```
+
+**Pie charts**
+
+```{python}
+#| echo: true
+x = [35, 20, 30, 40, 50, 30]
+y = ['Apple', 'Bananna', 'Grapes', 'Orange', 'PineApple', 'Dragon Fruit']
+plt.title('Pie Chart')
+plt.pie(x, labels=y)
+plt.show()
+
+```
+
 ### Seaborn
+
+Seaborn module for visualization
+
+**Line Graph**
+
+```{python}
+
+#| echo: true
+
+# importing packages
+import seaborn as sns
+# loading dataset
+data = sns.load_dataset("iris")
+# draw lineplot
+sns.lineplot(x="sepal_length", y="sepal_width", data=data)
+
+```
+
+**Scatter Graph**
+
+```{python}
+#| echo: true
+
+data = sns.load_dataset("iris")
+sns.scatterplot(data=data)
+```
 
@@
 
-## Folder Structure & File Management
+## Folder Structure & File Management {visibility="hidden"}
 
-# SQL
+# SQL {visibility="hidden"}
 
-## SQL Introduction  {.smaller}
+## SQL Introduction  {.smaller} {visibility="hidden"}

### 2024-11-11 MsaPythonDataAnalytics.qmd
'day1'

@@
 
+```{python}
+#| echo: true
+#| eval: false
+print(first_string.capitalize())
+print(first_string.title())
+print(first_string.upper())
+print(first_string.swapcase())
+print(first_string.find('is'))
+print(first_string.replace('first', 'second'))
+print(first_string.strip())
+print(first_string.isalnum())
+print(first_string.isalpha())
+print(first_string.isdigit())
+print(first_string.isprintable())
+```
+
 ***List***
@@
 
-### Iterations
+### Conditions & Iterations
+
+Conditions criteria where we compare values and decide what step to take.
+Example of conditional criteria `if-else` , `if-elif-else`, `while`.
+
+```{python}
+#| echo: true
+#| eval: false
+
+if condition:
+  #do something
+else:
+  #do something
+```
+
+Comparison operators go hand in hand with conditions.
+Comparison operators inculde `==` , `<=`, `>=`, `|`, `&`, `or`, `and` e.t.c
 
@@
 
+### Date & Time
+
+- **Date:** Handles dates without time.
+- **POSIXct:** Handles date & time in calendar time.
+- **POSIXlt:** Handles date & time in local time.
+- **Hms:** Parses periods with hour, minute, and second
+- **Timestamp:** Represents a single pandas date & time
+- **Interval:** Defines an open or closed range between dates and times
+- **Time delta:** Computes time difference between different datetimes
+
+***ISO8601 datetime format***  (YYYY-MM-DD HH:MM:SS TZ)
+
+```{python}
+#| echo: true
+#| eval: true
+
+import datetime as dt
+import time as tm
+import pytz
+import pandas as pd
+
+```
+
+```{python}
+#| echo: true
+#| eval: true
+
+# Get the current date
+dt.date.today()
+
+# Get the current date and time
+dt.datetime.now()
+```
+
+***Parsing dates, datetimes, and times***
+
+```{python}
+#| echo: true
+#| eval: true
+
+iso = pd.to_datetime(['2018-10-26 12:00', '2018-10-26 13:00'], utc=True)
+
+us = ['07/20/1969 20:17:40','11/19/1969 06:54:35']
+
+non_us = ['20/07/1969 20:17:40','19/11/1969 06:54:35']
+
+parts = pd.DataFrame({
+  'year' : [1969,1969,1971],
+  'month' : [7,11,2],
+  'day' : [20,19,5]
+})
+
+# Parse dates in ISO format
+pd.to_datetime(iso)
+
+# Parse dates in US format
+pd.to_datetime(us, dayfirst=False)
+
+# Parse dates in NON US format
+pd.to_datetime(non_us, dayfirst=True)
+
+# Parse dates, guessing a single format
+pd.to_datetime(iso, infer_datetime_format=True)
+
+# Parse dates in single, specified format
+pd.to_datetime(iso, format="%Y-%m-%d %H:%M:%S")
+
+# Parse dates in single, specified format
+pd.to_datetime(us, format="%m/%d/%Y %H:%M:%S")
+
+# Make dates from components
+pd.to_datetime(parts)
+```
+
+***Extracting components***
+
+```{python}
+#| echo: true
+#| eval: true
+
+# Parse strings to datetimes
+dttm = pd.to_datetime(iso)
+
+# Get year from datetime pandas series
+dttm.year
+
+# Get day of the year from datetime pandas series
+dttm.day_of_year
+
+# Get month name from datetime pandas series
+dttm.month_name()
+
+# Get day name from datetime pandas series
+dttm.day_name()
+
+# Get datetime.datetime format from datetime pandas series
+dttm.to_pydatetime()
+```
+
+**Rounding Dates**
+
+```{python}
+#| echo: true
+#| eval: true
+
+# Rounding dates to nearest time unit
+dttm.round('1min')
+
+# Flooring dates to nearest time unit
+dttm.floor('1min')
+
+# Ceiling dates to nearest time unit
+dttm.ceil('1min')
+```
+
+**Arithmetic**
+
+```{python}
+#| echo: true
+#| eval: true
+
+# Create two datetimes
+now = dt.datetime.now()
+then = pd.Timestamp('2021-09-15 10:03:30')
+
+# Get time elapsed as timedelta object
+now - then
+
+# Get time elapsed in seconds 
+(now - then).total_seconds()
+
+# Adding a day to a datetime
+dt.datetime(2022,8,5,11,13,50) + dt.timedelta(days=1)
+```
+
+**Time Intervals**
+
+```{python}
+#| echo: true
+#| eval: true
+
+# Create interval datetimes
+start_1 = pd.Timestamp('2021-10-21 03:02:10')
+finish_1 = pd.Timestamp('2022-09-15 10:03:30')
+start_2 = pd.Timestamp('2022-08-21 03:02:10')
+finish_2 = pd.Timestamp('2022-12-15 10:03:30')
+
+# Specify the interval between two datetimes
+pd.Interval(start_1, finish_1, closed='right')
+
+# Get the length of an interval
+pd.Interval(start_1, finish_1, closed='right').length
+
+# Determine if two intervals are intersecting
+pd.Interval(start_1, finish_1, closed='right').overlaps(pd.Interval(start_2, finish_2, closed='right'))
+```
+
+**Time Deltas**
+
+```{python}
+#| echo: true
+#| eval: true
+
+# Define a timedelta in days
+pd.Timedelta(7, "d")
+
+# Convert timedelta to seconds
+pd.Timedelta(7, "d").total_seconds()
+```
+
+## Joins & Merge {visibility="hidden"}
+
 :::
@@
 ## Python Visualization | Matplotlip & Seaborn
+
+Python provides so many modules for visualization but for purposes of the training we will deal with the two most common visualiation tools.
+
+1. **matplotlib**
+1. **seaborn**
+1. plotnine
+1. altair
+1. Bokeh
 
@@
 
+**Box plots**
+
+```{python}
+
+#| echo: true
+
+# Sample data
+data = {
+    'Category': ['A']*10 + ['B']*10,
+    'Value': [1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11]
+}
+
+df = pd.DataFrame(data)
+
+# Box plot
+df.boxplot(by='Category')
+plt.title('Box Plot Example')
+plt.suptitle('')
+plt.xlabel('Category')
+plt.ylabel('Value')
+plt.show()
+
+```
+
 ### Seaborn
@@
 
+```{python}
+#| echo: true
+
+#import seaborn as sns
+
+tips = sns.load_dataset("tips")
+
+sns.scatterplot(x="total_bill", y="tip", data=tips)
+```
+
+```{python}
+#| echo: true
+
+import seaborn as sns
+
+import matplotlib.pyplot as plt
+
+tips = sns.load_dataset("tips")
+
+# customize the scatter plot
+
+sns.scatterplot(x="total_bill", y="tip", hue="sex", size="size", sizes=(50, 200), data=tips)
+
+# add labels and title
+
+plt.xlabel("Total Bill")
+
+plt.ylabel("Tip")
+
+plt.title("Relationship between Total Bill and Tip")
+
+# display the plot
+
+plt.show()
+```
+
+### Pandas Inbuilt Visualization
+
+* df.plot.area     
+* df.plot.barh     
+* df.plot.density  
+* df.plot.hist     
+* df.plot.line     
+* df.plot.scatter
+* df.plot.bar      
+* df.plot.box      
+* df.plot.hexbin   
+* df.plot.kde      
+* df.plot.pie
+
 :::

### 2024-11-12 MsaPythonDataAnalytics.qmd
'update_day_4'

@@
 
+::: {.panel-tabset}
+
+### Intro
+
 ```{python}
@@
 
-```
-
-Array descriptors
-
-```{python}
-#| echo: true
-
-data.shape
-data.dtype
-data.ndim
-
-```
+a = np.array([1,2,3])
+b = np.array([(1.5,2,3), (4,5,6)], dtype = float)
+c = np.array([[(1.5,2,3), (4,5,6)],[(3,2,1), (4,5,6)]], dtype = float)
+
+np.zeros((3,4)) #Create an array of zeros
+np.ones((2,3,4),dtype=np.int16) #Create an array of ones
+d = np.arange(10,25,5)#Create an array of evenly spaced values (step value)
+np.linspace(0,2,9) #Create an array of evenlyspaced values (number of samples)
+e = np.full((2,2),7)#Create a constant array
+f = np.eye(2) #Create a 2X2 identity matrix
+np.random.random((2,2)) #Create an array with random values
+np.empty((3,2)) #Create an empty array
+
+```
+
+### Array descriptors & inspectors
+
+```{python}
+#| echo: true
+#| eval: false
+
+data.shape #Array dimensions
+len(a)#Length of array
+b.ndim #Number of array dimensions
+e.size #Number of array elements
+b.dtype  #Data type of array elements
+b.dtype.name  #Name of data type
+b.astype(int). #Convert an array to a different type
+
+```
+
+### Array Arithmetics
+
+```{python}
+#| echo: true
+#| eval: false
+
+g = a - b. #Subtraction
+
+np.subtract(a,b) #Subtraction
+b + a #Addition
+
+np.add(b,a) #Addition 
+a/b #Division
+
+np.divide(a,b) #Division 
+a * b #Multiplication
+
+np.multiply(a,b) #Multiplication 
+np.exp(b) #Exponentiation
+np.sqrt(b) #Square root
+np.sin(a)  #Print sines of an array
+np.cos(b) #Elementwise cosine
+np.log(a)#Elementwise natural logarithm
+e.dot(f) #Dot product
+
+```
+
+**Comparison**
+
+```{python}
+#| echo: true
+#| eval: false
+
+a == b #Elementwise comparison
+
+a< 2 #Elementwise comparison
+
+np.array_equal(a, b) #Arraywise comparison
+```
+
+**Sorting Arrays**
+
+```{python}
+#| echo: true
+#| eval: false
+
+a.sort() #Sort an array
+c.sort(axis=0) #Sort the elements of an array's axis
+```
+
+**Subsetting, slicing, indexing
+
+```{python}
+#| echo: true
+#| eval: false
+
+a[2] #Select the element at the 2nd index
+
+b[1,2] #Select the element at row 1 column 2(equivalent to b[1][2])
+
+a[0:2]#Select items at index 0 and 1
+
+b[0:2,1] #Select items at rows 0 and 1 in column 1
+
+b[:1] #Select all items at row0(equivalent to b[0:1, :])
+
+c[1,...] #Same as[1,:,:]
+
+a[ : : -1] #Reversed array a array([3, 2, 1])
+
+a[a<2] #Select elements from a less than 2
+
+b[[1,0,1, 0],[0,1, 2, 0]] #Select elements(1,0),(0,1),(1,2) and(0,0)
+
+b[[1,0,1, 0]][:,[0,1,2,0]] #Select a subset of the matrix’s rows and columns
+
+```
+
+### Array Manipulation
+
+```{python}
+#| echo: true
+#| eval: false
+
+i = np.transpose(b) #Permute array dimensions
+i.T #Permute array dimensions
+
+b.ravel() #Flatten the array
+g.reshape(3, -2) #Reshape, but don’t change data
+
+h.resize((2,6)) #Return a new arraywith shape(2,6)
+np.append(h,g) #Append items to an array
+np.insert(a,1,5)  #Insert items in an array
+np.delete(a,[1])  #Delete items from an array
+
+np.concatenate((a,d),axis=0) #Concatenate arrays
+
+np.vstack((a,b) #Stack arrays vertically(row wise)
+
+np.r_[e,f] #Stack arrays vertically(row wise)
+np.hstack((e,f)) #Stack arrays horizontally(column wise)
+
+np.column_stack((a,d)) #Create stacked column wise arrays
+
+np.c_[a,d] #Create stacked column wise arrays
+
+np.hsplit(a,3) #Split the array horizontally at the 3rd index
+
+np.vsplit(c,2) #Split the array vertically at the 2nd index
+
+```
+
+:::
 
@@
 Main difference of pandas with numpy is that it has indexed values and designed for working with tabular or heterogeneous data.
-
-::: {.panel-tabset}
-
-### Intro
 
@@
 
-## Joins & Merge {visibility="hidden"}
+### Joins & Merge
+
+**Merge***
+
+```{python}
+#| echo: true
+#| eval: false
+
+df1 = pd.DataFrame({'employee': ['Bob', 'Jake', 'Lisa', 'Sue'],
+                    'group': ['Accounting', 'Engineering', 'Engineering', 'HR']})
+df2 = pd.DataFrame({'employee': ['Lisa', 'Bob', 'Jake', 'Sue'],
+                    'hire_date': [2004, 2008, 2012, 2014]})
+
+df3 = pd.merge(df1, df2)
+df3
+
+df4 = pd.DataFrame({'group': ['Accounting', 'Engineering', 'HR'],
+                    'supervisor': ['Carly', 'Guido', 'Steve']})
+
+pd.merge(df3, df4)
+
+#Many-to-many joins
+
+df5 = pd.DataFrame({'group': ['Accounting', 'Accounting',
+                              'Engineering', 'Engineering', 'HR', 'HR'],
+                    'skills': ['math', 'spreadsheets', 'coding', 'linux',
+                               'spreadsheets', 'organization']})
+
+pd.merge(df1, df5)
+
+df3 = pd.DataFrame({'name': ['Bob', 'Jake', 'Lisa', 'Sue'],
+                    'salary': [70000, 80000, 120000, 90000]})
+
+pd.merge(df1, df3, left_on="employee", right_on="name")
+
+#Drop redundant column
+pd.merge(df1, df3, left_on="employee", right_on="name").drop('name', axis=1)
+
+```
+
+**Join**
+
+```{python}
+#| echo: true
+#| eval: false
+
+df1a = df1.set_index('employee')
+df2a = df2.set_index('employee')
+
+pd.merge(df1a, df2a, left_index=True, right_index=True)
+
+#For convenience, DataFrames implement the join() method, which performs a merge that defaults to joining on indices:
+
+df1a.join(df2a)
+
+df6 = pd.DataFrame({'name': ['Peter', 'Paul', 'Mary'],
+                    'food': ['fish', 'beans', 'bread']},
+                   columns=['name', 'food'])
+df7 = pd.DataFrame({'name': ['Mary', 'Joseph'],
+                    'drink': ['wine', 'beer']},
+                   columns=['name', 'drink'])
+
+pd.merge(df6, df7)
+
+#Inner Join
+
+pd.merge(df6, df7, how='inner')
+
+#Other options for the how keyword are 'outer', 'left', and 'right'. An outer join returns a join over the union of the input columns, and fills in all missing values with NAs
+
+pd.merge(df6, df7, how='outer')
+
+pd.merge(df6, df7, how='left')
+
+```
+
+### Let us get our hands dirty
+
+From the google drive link request for access and download the data [google drive](https://drive.google.com/drive/folders/1BkcXX95KMn_6b_Y8KwjAkGjzMtogWxd7?usp=sharing)
+
+1. Download the state data
+1. Merge the three different csv's
+1. Rank the population density for the different states in 2010
+  - population density calculation = population size / state size area
+1. Top 3 states
 

### 2024-11-12 raw/State_exercise.py

@@ -1,0 +1,52 @@
+pop = pd.read_csv('data/state-population.csv')
+areas = pd.read_csv('data/state-areas.csv')
+abbrevs = pd.read_csv('data/state-abbrevs.csv')
+
+
+merged = pd.merge(pop, abbrevs, how='outer',
+                  left_on='state/region', right_on='abbreviation')
+merged = merged.drop('abbreviation', 1) # drop duplicate info
+merged.head()
+
+
+merged.isnull().any()
+
+
+merged[merged['population'].isnull()].head()
+
+
+
+merged.loc[merged['state'].isnull(), 'state/region'].unique()
+
+
+merged.loc[merged['state/region'] == 'PR', 'state'] = 'Puerto Rico'
+merged.loc[merged['state/region'] == 'USA', 'state'] = 'United States'
+merged.isnull().any()
+
+
+
+final = pd.merge(merged, areas, on='state', how='left')
+final.head()
+
+
+final.isnull().any()
+
+final['state'][final['area (sq. mi)'].isnull()].unique()
+
+
+
+final.dropna(inplace=True)
+final.head()
+
+
+
+
+
+data2010.set_index('state', inplace=True)
+density = data2010['population'] / data2010['area (sq. mi)']
+
+
+density.sort_values(ascending=False, inplace=True)
+density.head()
+
+density.tail()

### 2024-11-14 MsaPythonDataAnalytics.qmd
'update3'

@@
 
+data.sort_values(by = 'k1') #sort
+
+```
+
+**Sample Cleaning**
+
+```{python}
+#| echo: true
+#| eval: false
+
+# Reading data using pandas
+df = pd.read_csv("https://rcs.bu.edu/examples/python/DataAnalysis/Salaries.csv")
+# List first 5 records
+df.head()
+
+#Select column
+
+df['sex']
+df.sex
+
+# #Group data using rank
+df_rank = df.groupby(["rank"])
+
+df_rank.head()
+
+# #Calculate mean value for each numeric column per each group
+df_rank.mean()
+
+# Once groupby object is created we can calculate various statistics for each group:
+
+#Calculate mean salary for each professor rank:
+df.groupby('rank')[['salary']].mean()
+
+# Note: If single brackets are used to specify the column (e.g. salary), then the output is Pandas Series object.
+# When double brackets are used the output is a Data Frame
+
+#Calculate mean salary for each professor rank:
+df.groupby(['rank'], sort=False)[['salary']].mean()
+
+# subset the rows in which the salary value is greater than $120K: 
+df_sub = df[df['salary'] > 120000]
+df_sub.head()
+
+#Select only those rows that contain female professors:
+df_f = df[df['sex'] == 'Female']
+
+#Selecting rows
+df[0:10]
+
+#Select rows by their labels:
+df_sub.loc[10:20,['rank','sex','salary']]
+
+#Select rows by their labels:
+df_sub.iloc[10:20,[0, 3, 4, 5]]
+
+#We can sort the data using 2 or more columns:
+df_sorted = df.sort_values( by =['service', 'salary'], ascending = [True, False])
+df_sorted.head(10)
+
+```
+
+**Sample 2***
+
+```{python}
+#| echo: true
+#| eval: false
+
+# Read a dataset with missing values
+flights = pd.read_csv("https://rcs.bu.edu/examples/python/DataAnalysis/flights.csv")
+
+# Select the rows that have at least one missing value
+flights[flights.isnull().any(axis=1)].head()
+
+#There are a number of methods to deal with missing values in the data frame:
+
+#df.dropna(): drop missing observations
+#df.dropna(how = "all"):  drop observtions where all cells is NA
+#df.dropna(axis = 1,how = "all"): drop column if all the values are missing
+#df.dropna(thresh = 5): Drop rows that contain less than 5 non-missing values
+#df.fillna(0): Replace missing values with zeros
+#df.isnull(): returns True if the value is missing
+#df.notnull(): Returns True for non-missing values
+
+```
+
+#### Aggregation Functions in Pandas
+
+Aggregation - computing a summary statistic about each group, i.e.
+compute group sums or means
+compute group sizes/counts
+
+Common aggregation functions:
+
+min, max
+count, sum, prod
+mean, median, mode, mad
+std, var
+
+```{python}
+#| echo: true
+#| eval: false
+
+flights[['dep_delay','arr_delay']].agg(['min','mean','max'])
+
+flights.groupby('origin').count()
+
+flights['origin'].value_counts()
+
 ```
@@
 1. Top 3 states
+
+### Bonus material
+
+**Collapsing data into categories**
+Map categories to fewer ones: reducing categories in categorical column.
+
+```{python}
+#| echo: true
+#| eval: false
+
+# Create mapping dictionary and replace
+mapping = {'Microsoft':'ComputerOS', 'MacOS':'ComputerOS', 'Linux':'ComputerOS',
+'IOS':'MobileOS', 'Android':'MobileOS'}
+devices['operating_system'] = devices['operating_system'].replace(mapping)
+devices['operating_system'].unique()
+
+# Replace "+" with "00"
+phones["Phone number"] = phones["Phone number"].str.replace("+", "00")
+phones
+
+# Replace "-" with nothing
+phones["Phone number"] = phones["Phone number"].str.replace("-", "")
+phones
+
+# Replace phone numbers with lower than 10 digits to NaN
+digits = phones['Phone number'].str.len()
+phones.loc[digits < 10, "Phone number"] = np.nan
+phones
+
+# Replace letters with nothing
+phones['Phone number'] = phones['Phone number'].str.replace(r'\D+', '')
+phones.head()
+
+# Treating date data
+birthdays['Birthday'] = birthdays['Birthday'].dt.strftime("%d-%m-%Y")
+birthdays.head()
+
+sum_classes = flights[['economy_class', 'business_class', 'first_class']].sum(axis = 1)
+passenger_equ = sum_classes == flights['total_passengers']
+# Find and filter out rows with inconsistent passengers
+inconsistent_pass = flights[~passenger_equ]
+consistent_pass = flights[passenger_equ]
+
+# Drop missing values
+airquality_dropped = airquality.dropna(subset = ['CO2'])
+airquality_dropped.head()
+
+#Replacing with statistical measures
+co2_mean = airquality['CO2'].mean()
+airquality_imputed = airquality.fillna({'CO2': co2_mean})
+airquality_imputed.head()
+
+# Multiple grouped summaries
+dogs.groupby("color")["weight_kg"].agg([min, max, sum])
+
+# Grouping by multiple variables
+dogs.groupby(["color", "breed"])["weight_kg"].mean()
+
+dogs.groupby(["color", "breed"])[["weight_kg", "height_cm"]].mean()
+
+# Group by to pivot table
+dogs.pivot_table(values="weight_kg", index="color")
+
+# Different statistics
+import numpy as np
+dogs.pivot_table(values="weight_kg", index="color", aggfunc=np.median)
+
+# Multiple statistics
+dogs.pivot_table(values="weight_kg", index="color", aggfunc=[np.mean, np.median])
+
+dogs.pivot_table(values="weight_kg", index="color", columns="breed")
+
+# Filling missing values in pivot tables
+dogs.pivot_table(values="weight_kg", index="color", columns="breed", 
+fill_value=0)
+
+# Summing with pivot tables
+dogs.pivot_table(values="weight_kg", index="color", columns="breed",
+fill_value=0, margins=True)
+
+```
 

### 2024-11-20 MsaPythonDataAnalytics.qmd
'update6'
'update7'

@@
     theme: serif
-    smaller: true
+    smaller: false
     scrollable: true
@@
 
+### Plotnine
+
+## {fullscreen=true}
+
+<iframe class="stretch" data-src="https://plotnine.org/reference/geom_point.html#basic-scatter-plot"></iframe>
+
 ### Pandas Inbuilt Visualization
@@
 
+# Bonus Material
+
+## {fullscreen=true}
+
+<iframe class="stretch" data-src="https://docs.google.com/presentation/d/16T9NAqRoLzOy1h5PtO_uoHN2lE037w9o/edit#slide=id.p1"></iframe>
+
+## {fullscreen=true}
+
+<iframe class="stretch" data-src="https://docs.google.com/presentation/d/1ONt4y1iNkKB0bEZXBqucG--xwOIcoE8R/edit#slide=id.p1"></iframe>
+
+## {fullscreen=true}
+
+<iframe class="stretch" data-src="https://pandas.pydata.org/pandas-docs/stable/user_guide/10min.html"></iframe>
+
 ## Folder Structure & File Management {visibility="hidden"}
